# Inference {#sec-inference}

This file uses posterior predictions to make inferences and evaluate hypotheses.

For more details on this kind of workflow for making inferences, I would suggest reading the following:

1. [Statistical rethinking](https://xcelab.net/rm/) by Richard McElreath [@mcelreathStatisticalRethinkingBayesian2020]. This provides a nice overview of prediction in the context of Bayesian data analysis.
2. [Model to meaning](https://marginaleffects.com/) by Vincent Arel-Bundock [@arel-bundockModelMeaningHow2025], as well as these related journal articles [@arel-bundockHowInterpretStatistical2024; @rohrerModelsPredictionMachines2026]. These articles provide the intuition and the tools to make it clear why it can often be advantageous to take a model and make predictions in a natural scale that that data were recorded in, rather than parameters in a regression model.

```{r}
#| label: setup
#| include: false
## load packages
pkgs <- c(
  "tidyverse",
  "patchwork",
  "ggdist",
  "brms",
  "bayesplot",
  "tidybayes",
  "parallel",
  "future",
  "furrr"
)
vapply(
  pkgs,
  library,
  logical(1),
  character.only = TRUE,
  logical.return = TRUE
)

## set options
## other settings
options(
  brms.backend = "cmdstanr",
  mc.cores = parallel::detectCores(),
  future.fork.enable = TRUE
  #future.rng.onMisuse = "ignore"
) ## automatically set in RStudio

supportsMulticore()

detectCores()

# plan(multicore)
## use multisession instead of multicore for Stan compatibility on Mac
plan(multisession, workers = 7) ##
# Set theme for all plots
theme_set(
  theme_bw() +
    theme(
      text = element_text(size = 18, face = "bold"),
      title = element_text(size = 18, face = "bold"),
      legend.position = "bottom"
    )
)
```

# 1. Read in the data and models

```{r}
#| label: read-data
#| code-fold: false
dat <- read_csv(
  "data/dat.csv",
  show_col_types = FALSE
) |>
  mutate(
    group = factor(
      group,
      levels = c("control", "embodied")
    )
  )
head(dat)
# glimpse(dat)
# summary(dat)
```

read models, if already computed

```{r}
#| label: read-models
#| eval: true # always runs — fast
#| code-fold: false
fit_b2 <- readRDS("models/fit_b2.rds")
summary(fit_b2)
```

# 2. Calculate posterior predictions

Here we try a simple version, which is the population-average probability for a hypothetical average participant on a hypothetical average item.

```{r}
#| label: calc-posterior-preds
#| code-fold: false 
nd <- tibble(
  group = c("control", "embodied"),
  pre_score_z = 0
)

# all draws per group
epred <- fit_b2 |>
  tidybayes::add_epred_draws(newdata = nd, re_formula = NA)

# difference (all draws)
epred_diff <- epred |>
  tidybayes::compare_levels(.epred, by = group)

# plot group epreds
plot_epred <- epred |>
  ggplot(aes(x = .epred, y = group, fill = group)) +
  stat_halfeye(alpha = 0.7) +
  labs(x = "P(correct)", y = NULL)

# plot the difference
plot_diff <- epred_diff |>
  ggplot(aes(x = .epred)) +
  stat_halfeye(alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Δ P(correct) [embodied - control]", y = NULL)

p1 <- plot_epred | plot_diff
p1

# summarise if needed
epred |>
  median_qi(.epred, .width = 0.95)

epred_diff |>
  median_qi(.epred, .width = 0.95)
```